---
title: "B201A HW2"
author: "Lillian Chen"
date: "10/27/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

1. A study investigating stopping distances (in feet) of automobiles traveling at different speeds yielded the following results:

Automobile speed	 n	   Mean	Std. Deviation
     10 m.p.h.		20	    20.2		       5.1
     20 m.p.h.		20	    36.3		       8.4	     
     30 m.p.h.		  5	    66.4		     15.5

(a)	Carry out a one-way analysis of variance (making the usual equal-variance assumption), and summarize whether the ANOVA F-test suggests that the hypothesis of equality of group means should be rejected at the $\alpha = 0.05$ level.
```{r}

```


(b)	Carry out two-sample t-tests comparing the 10-m.p.h. group to the 20-m.p.h. group assuming both equal and unequal variances.  Comment on similarities and differences in the findings, and in particular on the extent to which you think that possible violation of the equal-variance assumption affects your interpretations.

(c)	Carry out two-sample t-tests comparing the 20-m.p.h. group to the 30-m.p.h. group assuming both equal and unequal variances.  Comment on similarities and differences in the findings, and in particular on the extent to which you think that possible violation of the equal-variance assumption affects your interpretations.

2. The following data give the fastest running times (in seconds) recorded in the modern Olympic games through 1984 for both men and women at various distances.

Distance (m)				100	200	400	  800	1500	42195 (marathon)

X = Fastest time for women (sec) 	11.0	21.8	48.8	113.5	236.6	  8692.0

Y = Fastest time for men (sec)	 9.9	19.8	43.8	103.0	212.5	  7761.0

(a)	Calculate the correlation between X and Y based on the entire set of results.

(b)	Calculate the correlation between X and Y omitting the values for marathons.  Comment on the similarity or difference comparing your answer to the answer in part (a).

(c)	If you wanted to predict the fastest time for men and women over a distance of 1000 meters, would you be more inclined to rely on the entire sample or on the sample excluding the results from marathons?  Explain your reasoning.




3. (Glantz and Slinker problem 2.2):  Benzodiazepine tranquilizers (such as Valium) exert their physiological effects by binding to specific receptors in the brain.  This binding then interacts with a neurotransmitter, γ-amino butyric acid (GABA) to induce changes in nerve activity.  Because most direct methods of studying the effects of receptor binding are not appropriate for living human subjects, Hommer and coworkers (Archives of General Psychiatry, 1986, 43: 542-551) sought to study the effects of different doses of Valium on various readily measured physiological variables.  They then looked at the correlations among these variables to attempt to identify those that were most strongly linked to the effect of the drug.  Two of these variables were the sedation state $S$ induced by the drug and the blood level of the hormone cortisol $C$.  The data are presented below.

	Sedation score $S$		Cortisol $C$, $\mu$g/dl
		32				  6.6
		42				  7.4
		52				  8.8
		61				  9.7
		62				10.5
		65				11.8
		66				10.7

(a)  Calculate the sample mean and sample variance of both $S$ and $C$.

(b)  Draw (or produce on a computer) a scatter-plot relating $S$ and $C$.

(c)  Calculate the correlation between $S$ and $C$.

(d)  Using your judgment, comment on whether you think the association between $S$ and $C$ seems genuine in the sense that if you observed $S$ and $C$ on another 7 people it would likely yield a correlation in the same direction, or whether you think the estimated correlation based on the 7 different people might plausibly have a sign (i.e., positive or negative) in the opposite direction as the sign of correlation you calculated in part (c).

4. sConsider further the data from Problem 3.

	Sedation score $S$		Cortisol $C$, $\mu$g/dl
		32				  6.6
		42				  7.4
		52				  8.8
		61				  9.7
		62				10.5
		65				11.8
		66				10.7

(a)  For the regression of $S$ on $C$, calculate the least-squares estimate of the intercept, slope, and residual variance.

(b)  Calculate the standard error of the slope and the standard error of the intercept, and comment on whether each of these parameters is significantly different from 0.

(c)  Comment on what the findings from (a) and (b) tell you about whether the correlation is 
significantly different from 0.

